Weighted Model Counting/Integration from the Perspective of Boolean Algebras

### 2020-03-22 paper.tex
Statements for some new results

% Contributions:
% 1. A generalisation of WMC adapted to Boolean algebras.
% 2. We show that it is a valid measure.
% 3. We show that WMC assumes independence, i.e., cannot represent all measures.
% 4. We show how the BA can be extended with new 'literals' to represent any measure.
% 5. (Maybe) we establish a lower bound on the number of new literals, if we
% make no assumptions about independence.
% 6. (I guess I could) show how evidence = quotiening by a filter, and how
% quotiening by an ideal works
% 7. (WMI) If all weights are 1, we extend Boolean variables to discrete
% variables, and continuous variables are arbitrary subsets (in the
% sigma-algebra), then the WMI process calculates the probability of any event
% and doesn't require going down to the level of models (assuming a suitable
% probability measure).
% 8. If the weight of a literal (and its negation) is 1, then we don't need to
% look at its models--the probability can be calculated at any level. This
% extends to any number of variables.
% 9. How WMI transforms to our algebras. The boring version and what's wrong
% with it.

Previous/related work:
\begin{itemize}
\item Hailperin's approach to probability logic
  \cite{DBLP:journals/ndjfl/Hailperin84}
\item Nilsson's (somewhat successful) probabilistic logic
  \cite{DBLP:journals/ai/Nilsson86}
\item Semiring programming \cite{DBLP:journals/corr/BelleR16}
\item WMC with functions \cite{DBLP:conf/uai/Belle17}
\item WMI \cite{DBLP:conf/ijcai/BellePB15}
\item Measures on Boolean algebras: overview articles (from most cited to least
  cited)
  \begin{itemize}
  \item Horn and Tarski \cite{horn1948measures}
  \item Concerning measures on Boolean algebras \cite{gaifman1964concerning}
  \item Jech -- Measures on Boolean algebras (arXiv) \cite{jech2017measures}
  \end{itemize}
\item Measures on Boolean algebras: more specific articles
  \begin{itemize}
  \item On possibility and probability measures in finite Boolean algebras
    \cite{DBLP:journals/soco/CastineiraCT02}
  \item Representation of conditional probability measures
    \cite{krauss1968representation}
  \end{itemize}
\end{itemize}

@@

For clarity and succinctness, we will occasionally use three other operations
that can be defined using the original three\footnote{We use $+$ to denote
  symmetric difference because it is the additive operation of a Boolean ring.}:
\begin{align*}
  a \to b &= \neg a \lor b, \\
  a \leftrightarrow b &= (a \to b) \land (b \to a), \\
  a + b &= (a \land \neg b) \lor (\neg a \land b).
\end{align*}
We can also define a partial order $\le$ on $\mathbf{B}$ as $a \le b$ if $a = b
\land a$ (or, equivalently, $a \lor b = b$) for $a, b \in \mathbf{B}$.
Furthermore, let $a < b$ denote $a \le b$ and $a \ne b$. For the rest of this
paper, let $\mathbf{B}$ refer to the BA $(\mathbf{B}, \land, \lor, \neg, 0, 1)$.
For any $S \subseteq \mathbf{B}$, we write $\bigvee S$ for $\bigvee_{x \in S} x$
and call it the \emph{supremum} of $S$. Similarly, $\bigwedge S = \bigwedge_{x
  \in S} x$ is the \emph{infimum}. By convention, $\bigwedge \emptyset = 1$ and
$\bigvee \emptyset = 0$.

@@

@@

% TODO: Feedback: clarify the relationship between NWMC and the probability
% induced by a WMC formulation, i.e., Pr_{\Delta, w}(q) = \frac{WMC(\Delta \land
% q, w)}{WMC(\Delta, w)}.

\begin{proposition}
  $\WMC$ is a measure and $\nWMC$ is a probability measure.
\end{proposition}
\begin{proof}
  First, note that $\WMC$ is non-negative and $\WMC(0) = 0$ by definition. Next,
  let $x, y \in \mathbf{B}$ be such that $x \land y = 0$. We want to show that
  \begin{equation} \label{eq:additivity_proof}
    \WMC(x \lor y) = \WMC(x) + \WMC(y).
  \end{equation}
  If, say, $x = 0$, then \cref{eq:additivity_proof} becomes
  \[
    \WMC(y) = \WMC(0) + \WMC(y) = \WMC(y)
  \]
  (and likewise for $y = 0$). Thus we can assume that $x \ne 0 \ne y$ and use
  \cref{thm:representation} to write
  \[
    x = \bigvee_{i \in I} x_i \quad \text{and} \quad y = \bigvee_{j \in J} y_j
  \]
  for some sequences of atoms $(x_i)_{i \in I}$ and $(y_j)_{j \in J}$. If
  $x_{i'} = y_{j'}$ for some $i' \in I$ and $j' \in J$, then
  \[
    x \land y = \bigvee_{i \in I} \bigvee_{j \in J} x_i \land y_j = x_{i'} \land
    y_{j'} \ne 0,
  \]
  contradicting the assumption. This is enough to show that
  \begin{align*}
    \WMC(x \lor y) &= \WMC\left( \left( \bigvee_{i \in I} x_i \right) \lor \left(\bigvee_{j \in J} y_j \right) \right) = \sum_{i \in I} \WMC(x_i) + \sum_{j \in J} \WMC(y_j) \\
                   &= \WMC(x) + \WMC(y),
  \end{align*}
  finishing the proof that $\WMC$ is a measure. This immediately shows that
  $\nWMC$ is a probability measure since, by definition, $\nWMC(1) = 1$.
\end{proof}

@@

Let us note how our lower bound on the number of added literals compares to two
methods of translating a discrete probability distribution into a WMC problem
over a propositional knowledge base proposed by Darwiche
\cite{DBLP:conf/kr/Darwiche02} and Sang et al. \cite{sang2005solving}. Suppose
we have a discrete probability distribution with  $n$ variables, and the $i$-th
variable has $v_i$ values, for each $i \in [n]$. Interpreted as a logical
system, it has $\prod_{i=1}^n v_i$ models. My expansion would then use
\[
  \sum_{i=1}^n v_i + 2\prod_{i=1}^n v_i
\]
variables, i.e., a variable for each possible variable-value assignment, and two
additional variables for each model. Without making any independence
assumptions, the encoding by Darwiche \cite{DBLP:conf/kr/Darwiche02} would use
\[
  \sum_{i=1}^n v_i + \sum_{i=1}^n \prod_{j=1}^i v_j
\]
variables, while for the encoding by Sang et al. \cite{sang2005solving},
\[
  \sum_{i=1}^n v_i + \sum_{i=1}^n (v_i - 1) \prod_{j=1}^{i-1} v_j
\]
variables would suffice.

>>>>

% Contributions:
% 1. A generalisation of WMC adapted to Boolean algebras.
% 2. We show that it is a valid measure.
% 3. We show that WMC assumes independence, i.e., cannot represent all measures.
% 4. We show how the BA can be extended with new 'literals' to represent any measure.
% 5. (Maybe) we establish a lower bound on the number of new literals, if we
% make no assumptions about independence.
% 6. (I guess I could) show how evidence = quotiening by a filter, and how
% quotiening by an ideal works
% 7. (WMI) If all weights are 1, we extend Boolean variables to discrete
% variables, and continuous variables are arbitrary subsets (in the
% sigma-algebra), then the WMI process calculates the probability of any event
% and doesn't require going down to the level of models (assuming a suitable
% probability measure).
% 8. If the weight of a literal (and its negation) is 1, then we don't need to
% look at its models--the probability can be calculated at any level. This
% extends to any number of variables.
% 9. How WMI transforms to our algebras. The boring version and what's wrong
% with it.
% 10. Show how to transfer the measure to a quotient algebra and that the answer
% match with each other and with the original definition of WMC.

Previous/related work:
\begin{itemize}
\item Hailperin's approach to probability logic
  \cite{DBLP:journals/ndjfl/Hailperin84}
\item Nilsson's (somewhat successful) probabilistic logic
  \cite{DBLP:journals/ai/Nilsson86}
\item Semiring programming \cite{DBLP:journals/corr/BelleR16}
\item WMI \cite{DBLP:conf/ijcai/BellePB15}
\item Measures on Boolean algebras: overview articles (from most cited to least
  cited)
  \begin{itemize}
  \item Horn and Tarski \cite{horn1948measures}
  \item Concerning measures on Boolean algebras \cite{gaifman1964concerning}
  \item Jech -- Measures on Boolean algebras (arXiv) \cite{jech2017measures}
  \end{itemize}
\item Measures on Boolean algebras: more specific articles
  \begin{itemize}
  \item On possibility and probability measures in finite Boolean algebras
    \cite{DBLP:journals/soco/CastineiraCT02}
  \item Representation of conditional probability measures
    \cite{krauss1968representation}
  \end{itemize}
\end{itemize}

@@

For clarity and succinctness, we will occasionally use three other operations
that can be defined using the original three\footnote{We use $+$ to denote
  symmetric difference because it is the additive operation of a Boolean ring.}:
\begin{align*}
  a \to b &= \neg a \lor b, \\
  a \leftrightarrow b &= (a \land b) \lor (\neg a \land \neg b), \\
  a + b &= (a \land \neg b) \lor (\neg a \land b).
\end{align*}
We can also define a partial order $\le$ on $\mathbf{B}$ as $a \le b$ if $a = b
\land a$ (or, equivalently, $a \lor b = b$) for $a, b \in \mathbf{B}$.
Furthermore, let $a < b$ denote $a \le b$ and $a \ne b$. For the rest of this
paper, let $\mathbf{B}$ refer to the BA $(\mathbf{B}, \land, \lor, \neg, 0, 1)$.
For any $S \subseteq \mathbf{B}$, we write $\bigvee S$ for $\bigvee_{x \in S} x$
and call it the \emph{supremum} of $S$. Similarly, $\bigwedge S = \bigwedge_{x
  \in S} x$ is the \emph{infimum}. By convention, $\bigwedge \emptyset = 1$ and
$\bigvee \emptyset = 0$.

@@

% TODO: Start here. Define filters (ref the undergraduate book), filter
% generation (same ref, but read it first), and quotients.
\begin{definition}
\end{definition}

@@

\begin{proposition}
  $\WMC$ is a measure, and $\nWMC$ is a probability measure.
\end{proposition}
\begin{proof}
  First, note that $\WMC$ is non-negative and $\WMC(0) = 0$ by definition. Next,
  let $x, y \in \mathbf{B}$ be such that $x \land y = 0$. We want to show that
  \begin{equation} \label{eq:additivity_proof}
    \WMC(x \lor y) = \WMC(x) + \WMC(y).
  \end{equation}
  If, say, $x = 0$, then \cref{eq:additivity_proof} becomes
  \[
    \WMC(y) = \WMC(0) + \WMC(y) = \WMC(y)
  \]
  (and likewise for $y = 0$). Thus we can assume that $x \ne 0 \ne y$ and use
  \cref{thm:representation} to write
  \[
    x = \bigvee_{i \in I} x_i \quad \text{and} \quad y = \bigvee_{j \in J} y_j
  \]
  for some sequences of atoms $(x_i)_{i \in I}$ and $(y_j)_{j \in J}$. If
  $x_{i'} = y_{j'}$ for some $i' \in I$ and $j' \in J$, then
  \[
    x \land y = \bigvee_{i \in I} \bigvee_{j \in J} x_i \land y_j = x_{i'} \land
    y_{j'} \ne 0,
  \]
  contradicting the assumption. This is enough to show that
  \begin{align*}
    \WMC(x \lor y) &= \WMC\left( \left( \bigvee_{i \in I} x_i \right) \lor \left(\bigvee_{j \in J} y_j \right) \right) = \sum_{i \in I} \WMC(x_i) + \sum_{j \in J} \WMC(y_j) \\
                   &= \WMC(x) + \WMC(y),
  \end{align*}
  finishing the proof that $\WMC$ is a measure. This immediately shows that
  $\nWMC$ is a probability measure since, by definition, $\nWMC(1) = 1$.
\end{proof}

Given a theory $\Delta$ in a logic $\mathcal{L}$, the usual way of using WMC to
compute the probability of a query $q$ is
\cite{DBLP:conf/uai/Belle17,DBLP:conf/aaai/SangBK05}
\[
  \Pr_{\Delta, w}(q) = \frac{\WMC_w(\Delta \land q)}{\WMC_w(\Delta)}.
\]
In our algebraic formulation, this can be computed in two different ways:
\begin{itemize}
\item as $\frac{\WMC_w(\Delta \land q)}{\WMC_w(\Delta)}$ in $B(\mathcal{L})$,
\item and as $\nWMC_w([q])$ in $B(\Delta)$.
\end{itemize}
But how does the measure defined on $B(\mathcal{L})$ transfer to $B(\Delta)$?

\begin{lemma}
  Let $\mathbf{B}$ be a BA, and let $F$ be a filter in $\mathbf{B}$ generated by
  $p \in \mathbf{B}$. Then, for any $a, b \in \mathbf{B}$, if $a/F = b/F$, then
  $a \land p = b \land p$.
\end{lemma}
\begin{proof}
  % TODO: reference for a/F = b/F iff a <-> b \in F
  % TODO: reference for the definition of a filter
\end{proof}

\begin{definition}[Measures on quotients] \label{def:measures_on_quotients}
  Let $\mathbf{B}$ be a BA with a measure $m : \mathbf{B} \to \mathbb{R}_{\ge
    0}$, and let $F$ be a filter in $\mathbf{B}$ generated by some $p \in
  \mathbf{B}$. We can define a measure $m^*$ on $\mathbf{B}/F$ as
  \[
    m^*([a]) = m(a \land p).
  \]
\end{definition}

\begin{proposition}
  The function $m^*$ in \cref{def:measures_on_quotients} is well-defined.
  Furthermore, if $m$ is a WMC measure, then so is $m^*$.
\end{proposition}
\begin{proof}
\end{proof}
% TODO: prove that the two ways to compute the probability give the same answer.

% Feedback: Clarify the relationship between $\nWMC$ and the probability induced
% by a WMC formulation.

@@

Let us note how our lower bound on the number of added literals compares to two
methods of translating a discrete probability distribution into a WMC problem
over a propositional knowledge base proposed by Darwiche
\cite{DBLP:conf/kr/Darwiche02} and Sang et al. \cite{DBLP:conf/aaai/SangBK05}.
Suppose we have a discrete probability distribution with  $n$ variables, and the
$i$th variable has $v_i$ values, for each $i \in [n]$. Interpreted as a logical
system, it has $\prod_{i=1}^n v_i$ models. My expansion would then use
\[
  \sum_{i=1}^n v_i + 2\prod_{i=1}^n v_i
\]
variables, i.e., a variable for each possible variable-value assignment, and two
additional variables for each model. Without making any independence
assumptions, the encoding by Darwiche \cite{DBLP:conf/kr/Darwiche02} would use
\[
  \sum_{i=1}^n v_i + \sum_{i=1}^n \prod_{j=1}^i v_j
\]
variables, while for the encoding by Sang et al. \cite{DBLP:conf/aaai/SangBK05},
\[
  \sum_{i=1}^n v_i + \sum_{i=1}^n (v_i - 1) \prod_{j=1}^{i-1} v_j
\]
variables would suffice.
